Computing {\em the least commitment} action model from state observations

### 2018-11-26 doc/ijcai-19-conformant/conformant-learning.tex
first conformant compilation version ijcai19 paper

In addition, the paper presets a new method to compute {\em the least commitment} action model from state observations that compiles this task into {\em conformant planning}.

@@

\section{Learning {\em the least commitment} action model from observations}

\subsection{The {\em least commitment} action model}
Now we are ready to formalize the learning task that we are addressing in this paper. The task of computing {\em the least commitment} action model is defined as a tuple $\tup{P,M,\mathcal{O}}$ where:

\begin{itemize}
\item $P=\tup{F,A[\cdot],I,G}$ is a planning problem where $A[\cdot]$ is a set of actions. For each $a\in A[\cdot]$, the semantics of $a$ is unknown; i.e. the functions $\rho$ and/or $\theta$ of $a$ are undefined.
\item $\mathcal{O}(\tau)$ is an observation of a trajectory $\tau(\pi,P)$ produced by the execution of an unknown plan $\pi$ that solves the planning problem $P$.
\item $M$ is the set of different planning models for the actions in $A[\cdot]$. A model $\mathcal{M}\in M$ defines the semantics of every action in $A[\cdot]$. Planning models differ in the $\tup{\rho,\theta}$ functions of the actions but they all use the same set of state variables $F$.
\end{itemize}

The {\em solution} to this task is {\em the least commitment} action model which is defined as the minimum subset of models $M^*\in M$ that is compliant with the input observation $\mathcal{O}(\tau)$.

\subsection{The hypothesis space}
\strips\ action schemata provide a compact representation specifying action models. {\em A \strips\ action schema} $\xi$ is defined by four lists: A list of {\em parameters} $pars(\xi)$, and three list of predicates (namely $pre(\xi)$, $del(\xi)$ and $add(\xi)$) that shape the kind of fluents that can appear in the {\em preconditions}, {\em negative effects} and {\em positive effects} of the actions induced from that schema.

@@

\item {\em Domain-specific constraints}. One can introduce domain-specific knowledge to constrain further the space of possible schemata. For instance, in the {\em blocksworld} one can claim that {\small\tt on($v_1$,$v_1$)} and {\small\tt on($v_2$,$v_2$)} will not appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of an action schema $\xi$ because of the semantic of the {\tt\small on} predicate (i.e. only one block can be on top of another block). 
\end{enumerate}

@@

\begin{enumerate}
\item[3.] {\em Observation constraints}. An observation of the execution of a plan further constraints the space of possible action schemata. This \emph{semantic knowledge} included in the observations introduce a third type of constraints and that can also be added to the set $\mathcal{C}$.
\end{enumerate}

\subsection{Learning {\em the least commitment} model with conformant planning}
Given a \strips\ action schema $\xi$, a propositional encoding for the {\em preconditions}, {\em negative} and {\em positive} effects of that schema can be represented with fluents of the kind $[pre|del|add]\_e\_\xi$ such that $e\in{\mathcal I}_{\Psi,\xi}$ is a single element from the set of interpretations of predicates $\Psi$ over the corresponding variable names $\Omega_\xi$. Figure~\ref{fig:encoding} shows the propositional encoding for the six action schema defined in Figure~\ref{fig:model-example}.

The interest of having a propositional encoding for \strips\ action schema is that, using {\em conditional effects}, it allows to compactly define {\em editable actions}. Actions whose semantics is given by the value of the $[pre|del|add]\_e\_\xi$ fluents at the current state. Given an operator schema $\xi\in\mathcal{M}$ its {\em editable} version is formalized as:
\begin{small}  
\begin{align*}
\hspace*{7pt}\pre(\mathsf{editable_{\xi}})=&\{pre\_e\_\xi\implies e\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}\\
\cond(\mathsf{editable_{\xi}})=&\{del\_e\_\xi\}\rhd\{\neg e\}_{\forall e\in{\mathcal I}_{\Psi,\xi}},\\
&\{add\_e\_\xi\}\}\rhd\{e\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}.
\end{align*}
\end{small}

Figure~\ref{fig:editable} shows the PDDL encoding of the {\em editable} {\tt\small inc-x(?v1,?v2)} schema for robot navigation in a $n\times n$ grid (Figure~\ref{fig:model-example}). Note that this editable schema, when the fluents of Figure~\ref{fig:encoding} hold, behaves exactly as defined in Figure~\ref{fig:model-example}.

\begin{figure}
  \begin{tiny}  
  \begin{verbatim}
(:action editable_inc-x
  :parameters (?v1 ?v2)
  :precondition
    (and (or (not (pre_xcoord_v1_inc-x)) (xcoord ?v1))
         (or (not (pre_xcoord_v2_inc-x)) (xcoord ?v2))
         (or (not (pre_ycoord_v1_inc-x)) (xcoord ?v1))                       
         (or (not (pre_ycoord_v2_inc-x)) (xcoord ?v2))
         (or (not (pre_q0__inc-x)) (q0))
         (or (not (pre_q1__inc-x)) (q1)))
         (or (not (pre_next_v1_v1_inc-x)) (next ?v1 ?v1)))
         (or (not (pre_next_v1_v2_inc-x)) (next ?v1 ?v2)))
         (or (not (pre_next_v2_v1_inc-x)) (next ?v2 ?v1)))
         (or (not (pre_next_v2_v2_inc-x)) (next ?v2 ?v2))))
    :effect (and
       (when (del_xcoord_v1_inc-x) (not (xcoord ?v1)))
       (when (del_xcoord_v2_inc-x) (not (xcoord ?v2)))
       (when (del_ycoord_v1_inc-x) (not (xcoord ?v1)))
       (when (del_ycoord_v2_inc-x) (not (xcoord ?v2)))
       (when (del_q0__inc-x) (not (q0)))
       (when (del_q1__inc-x) (not (q1)))
       (when (del_next_v1_v1_inc-x) (not (next ?v1 ?v1)))
       (when (del_next_v1_v2_inc-x) (not (next ?v1 ?v2)))
       (when (del_next_v2_v1_inc-x) (not (next ?v2 ?v1)))
       (when (del_next_v2_v2_inc-x) (not (next ?v2 ?v2)))

       (when (add_xcoord_v1_inc-x) (xcoord ?v1))
       (when (add_xcoord_v2_inc-x) (xcoord ?v2))
       (when (add_ycoord_v1_inc-x) (xcoord ?v1))
       (when (add_ycoord_v2_inc-x) (xcoord ?v2))
       (when (add_q0__inc-x) (q0))
       (when (add_q1__inc-x) (q1))
       (when (add_next_v1_v1_inc-x) (next ?v1 ?v1))
       (when (add_next_v1_v2_inc-x) (next ?v1 ?v2))
       (when (add_next_v2_v1_inc-x) (next ?v2 ?v1))
       (when (add_next_v2_v2_inc-x) (next ?v2 ?v2)))
  \end{verbatim}           
  \end{tiny}  
 \caption{\small Editable version of the {\tt\small inc-x(?v1,?v2)} schema for robot navigation in a $n\times n$ grid.}
\label{fig:editable}
\end{figure}

The fluents $\{test_j\}_{1\leq j\leq m}$, indicating the state observation $s_j\in\mathcal{O}(\tau)$ where the action model is validated. The new goals $G'=G\bigcup\{at_n,test_m\}$ and the new actions for {\em validating} the edited models at the $s_j$ observed states, {\tt\small $0\leq j< m$}.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{validate_{j}})=&s_j\cup\{test_{j-1}\},\\
\cond(\mathsf{validate_{j}})=&\{\emptyset\}\rhd\{\neg test_{j-1}, test_j,\\
                            &\{mode_{edit}\}\rhd\{\neg mode_{edit}, mode_{val}\}.
\end{align*}
\end{small}

>>>>

In addition, the paper presets a new method to compute {\em the least commitment} action model from state observations. The method assumes that action models are specified as \strips\ action schema and builds on top of off-the-shelf {\em conformant planning} algorithms.

@@

\section{Learning {\em the least commitment} action model from state observations}
Now we are ready to formalize the learning of the {\em the least commitment} action model from state observations and our approach to address it via {\em conformant planning}.

\subsection{The {\em least commitment} action model}
The task of computing {\em the least commitment} action model is defined as a tuple $\tup{P,M,\mathcal{O}}$ where:

\begin{itemize}
\item $P=\tup{F,A[\cdot],I,G[\cdot]}$ is a planning problem where $A[\cdot]$ is a set of actions. For each $a\in A[\cdot]$, the semantics of $a$ is unknown; i.e. the functions $\rho$ and/or $\theta$ of $a$ are undefined. The set of goal conditions $G[\cdot]$ is also unknown.
\item $M$ is the set of different planning models for the actions in $A[\cdot]$. A model $\mathcal{M}\in M$ defines the semantics of every action in $A[\cdot]$. Planning models differ in the $\tup{\rho,\theta}$ functions of the actions but they all use the same set of state variables $F$.
\item $\mathcal{O}(\tau)$ is a sequence of state observations coming from the observation of a trajectory $\tau(\pi,s_0)$ produced by the execution of certain unknown plan $\pi$.
\end{itemize}

The {\em solution} to this task is {\em the least commitment} action model which defines the smallest subset of models $M^*\subseteq M$ that is {\em conformant} with the input observation $\mathcal{O}(\tau)$.

\subsection{The space of possible action models}
This work focuses on the lerning of action models that are specified as \strips\ action schema. \strips\ action schemata provide a compact representation for specifying action models. {\em A \strips\ action schema} $\xi$ is defined by four lists: A list of {\em parameters} $pars(\xi)$, and three list of predicates (namely $pre(\xi)$, $del(\xi)$ and $add(\xi)$) that shape the kind of fluents that can appear in the {\em preconditions}, {\em negative effects} and {\em positive effects} of the actions induced from that schema.

@@

\item {\em Domain-specific constraints}. One can introduce domain-specific knowledge to constrain further the space of possible schemata. For instance, in the {\em blocksworld} one can argue that {\small\tt on($v_1$,$v_1$)} and {\small\tt on($v_2$,$v_2$)} will not appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists of an action schema $\xi$ because of the semantic of the {\tt\small on} predicate (i.e. only one block can be on top of another block). 
\end{enumerate}

@@

\begin{enumerate}
\item[3.] {\em Observation constraints}. A sequence of state observations $\mathcal{O}(\tau)$ constraints further the space of possible action schemata. This \emph{semantic knowledge} included in the observations introduce a third type of constraints and can also be added to the set $\mathcal{C}$.
\end{enumerate}

\subsection{Learning {\em the least commitment} model with conformant planning}
Given a classical planning problem $P=\tup{F,A[\cdot],I,G[\cdot]}$ and a sequence of state observations $\mathcal{O}(\tau)$ coming from the observation of a $\tau(\pi,I)$ trajectory, we show here that we can build a {\em conformant planning problem} $P_c$ in linear time and space and whose solution is a plan that induces the {\em least commitment} model that is conformant with $\mathcal{O}(\tau)$.

In more detail, our compilation defines a {\em conformant planning problem} $P_c=\tup{F_c,A_c,\Upsilon,G}$ such that:
\begin{itemize}
\item The set of fluents $F_c$ extends $F$ with two new sets of fluents:
      \begin{itemize}
      \item $\{test_j\}_{1\leq j\leq m}$, indicating the state observation $s_j\in\mathcal{O}(\tau)$ where the action model is validated
      \item The fluents $[pre|del|add]\_e\_\xi$ such that $e\in{\mathcal I}_{\Psi,\xi}$ for a propositional encoding of the {\em preconditions}, {\em negative} and {\em positive} effects of schema $\xi$. Figure~\ref{fig:propositional} shows the propositional encoding of the unstack action from the blocksworld.
      \end {itemize}

\begin{figure}
  \begin{tiny}  
  \begin{verbatim}
(pre_holding_v1_stack) (pre_clear_v2_stack)
(del_holding_v1_stack) (del_clear_v2_stack)
(add_clear_v1_stack) (add_handempty_stack) (add_on_v1_v2_stack)
  \end{verbatim}           
  \end{tiny}  
 \caption{\small Propositional encoding of the {\tt\small stack(?v1,?v2)} schema.}
\label{fig:propositional}
\end{figure}

      \item The set of actions $A_c$ contains now actions of two different kinds:
\begin{itemize}
      \item Actions for {\em validating} the edited models at the $s_j$ observed states, {\tt\small $0\leq j< m$}
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{validate_{j}})=&s_j\cup\{test_{j-1}\},\\
\cond(\mathsf{validate_{j}})=&\{\emptyset\}\rhd\{\neg test_{j-1}, test_j,\\
                            &\{mode_{edit}\}\rhd\{\neg mode_{edit}, mode_{val}\}.
\end{align*}
\end{small}

      \item Actions whose semantics is given by the value of the $[pre|del|add]\_e\_\xi$ fluents at the current state. Given an operator schema $\xi\in\mathcal{M}$ its {\em editable} version is formalized as:
\begin{small}  
\begin{align*}
\hspace*{7pt}\pre(\mathsf{editable_{\xi}})=&\{pre\_e\_\xi\implies e\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}\\
\cond(\mathsf{editable_{\xi}})=&\{del\_e\_\xi\}\rhd\{\neg e\}_{\forall e\in{\mathcal I}_{\Psi,\xi}},\\
&\{add\_e\_\xi\}\}\rhd\{e\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}.
\end{align*}
\end{small}
Figure~\ref{fig:editable} shows the PDDL encoding of the {\em editable} {\tt\small inc-x(?v1,?v2)} schema for robot navigation in a $n\times n$ grid (Figure~\ref{fig:model-example}). Note that this editable schema, when the fluents of Figure~\ref{fig:encoding} hold, behaves exactly as defined in Figure~\ref{fig:model-example}.     
      \end{itemize}

\item The new goals are $G_c=\{test_m\}$ .
\end{itemize}

\begin{figure}
  \begin{tiny}  
  \begin{verbatim}
(:action stack
  :parameters (?o1 - object ?o2 - object)
  :precondition
   (and (or (not (pre_on_v1_v1_stack)) (on ?o1 ?o1))
        (or (not (pre_on_v1_v2_stack)) (on ?o1 ?o2))
        (or (not (pre_on_v2_v1_stack)) (on ?o2 ?o1))
        (or (not (pre_on_v2_v2_stack)) (on ?o2 ?o2))
        (or (not (pre_ontable_v1_stack)) (ontable ?o1))
        (or (not (pre_ontable_v2_stack)) (ontable ?o2))
        (or (not (pre_clear_v1_stack)) (clear ?o1))
        (or (not (pre_clear_v2_stack)) (clear ?o2))
        (or (not (pre_holding_v1_stack)) (holding ?o1))
        (or (not (pre_holding_v2_stack)) (holding ?o2))
        (or (not (pre_handempty_stack)) (handempty)))
  :effect
   (and (when (del_on_v1_v1_stack) (not (on ?o1 ?o1)))
        (when (del_on_v1_v2_stack) (not (on ?o1 ?o2)))
        (when (del_on_v2_v1_stack) (not (on ?o2 ?o1)))
        (when (del_on_v2_v2_stack) (not (on ?o2 ?o2)))
        (when (del_ontable_v1_stack) (not (ontable ?o1)))
        (when (del_ontable_v2_stack) (not (ontable ?o2)))
        (when (del_clear_v1_stack) (not (clear ?o1)))
        (when (del_clear_v2_stack) (not (clear ?o2)))
        (when (del_holding_v1_stack) (not (holding ?o1)))
        (when (del_holding_v2_stack) (not (holding ?o2)))
        (when (del_handempty_stack) (not (handempty)))
        (when (add_on_v1_v1_stack) (on ?o1 ?o1))
        (when (add_on_v1_v2_stack) (on ?o1 ?o2))
        (when (add_on_v2_v1_stack) (on ?o2 ?o1))
        (when (add_on_v2_v2_stack) (on ?o2 ?o2))
        (when (add_ontable_v1_stack) (ontable ?o1))
        (when (add_ontable_v2_stack) (ontable ?o2))
        (when (add_clear_v1_stack) (clear ?o1))
        (when (add_clear_v2_stack) (clear ?o2))
        (when (add_holding_v1_stack) (holding ?o1))
        (when (add_holding_v2_stack) (holding ?o2))
        (when (add_handempty_stack) (handempty))))
  \end{verbatim}           
  \end{tiny}  
 \caption{\small Editable version of the {\tt\small stack(?v1,?v2)} schema.}
\label{fig:editable}
\end{figure}

Given a plan $\pi$ that solves the {\em conformant planning} problem $P_c$ that results from our compilation then, the {\em least commitment} model that is conformant with $\mathcal{O}(\tau)$ is extracted in linear time from the last state reached by $\pi$.

\subsection{Compilation properties}
